---
title: "Premier League"
author: "Asad Adnan"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: yes
    mode: selfcontained
---

## Introduction

I intend to understand how, Leicester City FC, was able to win the Premier League title despite being promoted to it only 2 years ago.

### Primary Questions
What was Leicester City FC doing so special that made them achieve this great feat?

## Data

The data that I am using for this project can be accessed from [Kaggle](https://www.kaggle.com/datasets/saife245/english-premier-league) 

```{r}
full_premier_league <- read.csv("C:/Users/asada/Downloads/archive (7)/final_dataset.csv")

premier_league1 <- read.csv("C:/Users/asada/Downloads/archive (7)/Datasets/2014-15.csv")

premier_league_14 <- read.csv("C:/Users/asada/Downloads/archive (7)/Datasets/2014-15.csv")

premier_league_13 <- read.csv("C:/Users/asada/Downloads/archive (7)/Datasets/2013-14.csv")

premier_league_12 <- read.csv("C:/Users/asada/Downloads/archive (7)/Datasets/2012-13.csv")


premier_league <- read.csv("C:/Users/asada/Downloads/archive (7)/Datasets/2015-16.csv")

```

There is never a reason for you to print an entire data frame within a document. If you want to provide something that can show your data, you can use the `paged_table` function:

```{r}
str(premier_league)

##Leicester wins, losses and draws throughout the season
leicester_home <- premier_league[premier_league$HomeTeam == "Leicester", ]

nrow(leicester_home[leicester_home$FTR== "H", ]) #12 wins

nrow(leicester_home[leicester_home$FTR== "D", ]) #6 draws

nrow(leicester_home[leicester_home$FTR== "A", ]) #1 losses

leicester_away <- premier_league[premier_league$AwayTeam == "Leicester", ]

nrow(leicester_away[leicester_away$FTR== "H", ]) #2 losses

nrow(leicester_away[leicester_away$FTR== "D", ]) #6 draws

nrow(leicester_away[leicester_away$FTR== "A", ]) #11 wins

##making Leicester bet data frame and statistics data frame
leicester <- premier_league[premier_league$HomeTeam == "Leicester" | premier_league$AwayTeam == "Leicester", ]

leicester_bet <- leicester[ , c(1:4,24:62)]

leicester_stats <- leicester[ , c(1:23)]

## average stats for all teams at home and away

avg_home_stats <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ HomeTeam, 
                            data = premier_league, 
                            FUN = mean)

avg_away_stats <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ AwayTeam, 
                            data = premier_league, 
                            FUN = mean)

## graphs

library(ggplot2)

ggplot(avg_home_stats, aes(FTHG, HS, color=HomeTeam)) +
  geom_point() +
  theme_minimal()

cor(avg_home_stats$FTHG, avg_home_stats$HS)

ggplot(avg_home_stats, aes(FTAG, AS, color=HomeTeam)) +
  geom_point() +
  theme_minimal()

cor(avg_home_stats$AS, avg_home_stats$FTAG)


## leicester wins

leicester_wins <- rbind(leicester_home[leicester_home$FTR== "H", ], leicester_away[leicester_away$FTR== "A", ])

leicester_home$Date <- as.Date(leicester_home$Date, 
                         format = "%d/%m/%Y")

leicester_away$Date <- as.Date(leicester_away$Date, 
                               format = "%d/%m/%Y")

leicester_wins <- leicester_wins[order(leicester_wins$Date), ]

##aggregated by full time results both home and away
aggregated_stats_h <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ HomeTeam + FTR, 
                              data = premier_league, 
                              FUN = mean)
aggregated_stats_a <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ AwayTeam + FTR, 
                            data = premier_league, 
                            FUN = mean)
aggregated_home_wins <- aggregated_stats_h[aggregated_stats_h$FTR == "H", ]
aggregated_away_wins <- aggregated_stats_a[aggregated_stats_a$FTR == "A", ]

## graphs

ggplot(aggregated_home_wins, aes(HomeTeam, HF)) +
  geom_col() +
  theme_minimal()

aggregated_home_wins1 <- aggregated_home_wins[aggregated_home_wins$HomeTeam == "Man City" | aggregated_home_wins$HomeTeam== "Leicester" | aggregated_home_wins$HomeTeam == "Man United" | aggregated_home_wins$HomeTeam == "Chelsea", ]

ggplot(aggregated_home_wins1, aes(FTAG, AS, color=HomeTeam))+
  geom_point()+
  theme_minimal()

avg_home_stats_lim <- avg_home_stats[avg_home_stats$HomeTeam == "Leicester" | avg_home_stats$HomeTeam == "Arsenal" | avg_home_stats$HomeTeam == "Tottenham" | avg_home_stats$HomeTeam == "Man City" | avg_home_stats$HomeTeam == "Man United", ]

ggplot(avg_home_stats_lim, aes(FTAG, AS, color=HomeTeam))+
  geom_point()+
  theme_minimal()

avg_away_stats_lim <- avg_away_stats[avg_away_stats$AwayTeam == "Leicester" | avg_away_stats$AwayTeam == "Man City" | avg_away_stats$AwayTeam == "Man United" | avg_away_stats$AwayTeam == "Chelsea", ]

ggplot(avg_away_stats_lim, aes(FTHG, HST, color=AwayTeam))+
  geom_point()+
  theme_minimal()

library(ggplot2)

ggplot(avg_home_stats_lim, aes(x = FTAG, y = AS, color = HomeTeam)) + 
  geom_point(size = 5) +  
  labs(title = "Home Games: Shots vs Goals Conceded",
       x = "Average Goals Conceded", y = "Average Shots", 
       color = "Home Team") +  
  theme_minimal() +  
  theme(
    text = element_text(size = 12),  # Increase font size for better readability
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels if necessary
    legend.position = "right"  # Adjust legend position
  ) +
  scale_color_manual(values = c("Arsenal" = "red", 
                                "Leicester" = "blue", 
                                "Tottenham" = "black",
                                "Man United" = "green",
                                "Man City" = "purple"))  # Set distinct colors for teams

```



```{r}
## goal difference in home games
aggregated_home_wins <- as.numeric(FTHG)
colnames(aggregated_home_wins)
aggregated_home_wins$gd <- aggregated_home_wins$FTHG - aggregated_home_wins$FTAG
aggregated_home_wins <- aggregated_home_wins[order(aggregated_home_wins$gd, decreasing =T), ]
head(aggregated_home_wins, 10)

winning_margins <- data.frame(aggregated_home_wins$HomeTeam, aggregated_home_wins$gd)
colnames(winning_margins)[1] <- "HomeTeam"
colnames(winning_margins)[2] <- "Average Goal Difference"

##goal difference in away games
aggregated_home_wins <- as.numeric(FTHG)
colnames(aggregated_home_wins)
aggregated_away_wins$gd <- aggregated_away_wins$FTAG - aggregated_away_wins$FTHG
aggregated_away_wins <- aggregated_away_wins[order(aggregated_away_wins$gd, decreasing =T), ]
head(aggregated_away_wins, 10)
View(winning_margins_away)
winning_margins_away <- data.frame(aggregated_away_wins$AwayTeam, aggregated_away_wins$gd)
colnames(winning_margins_away)[1] <- "AwayTeam"
colnames(winning_margins_away)[2] <- "Average Goal Difference"

#away games comparison

avg_away_stats <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ AwayTeam, 
                            data = premier_league, 
                            FUN = mean)

avg_away_stats_lim <- avg_away_stats[avg_away_stats$AwayTeam == "Leicester" |
                                       avg_away_stats$AwayTeam == "Man City" |
                                       avg_away_stats$AwayTeam == "Man United" |
                                       avg_away_stats$AwayTeam == "Arsenal" |
                                       avg_away_stats$AwayTeam == "Tottenham", ]

ggplot(avg_away_stats_lim, aes(x = FTAG, y = AST, color = AwayTeam)) + 
  geom_point(size = 5) +  
  labs(title = "Away Games: Shots vs Goals Scored",
       x = "Average Goals Scored", y = "Average Shots", 
       color = "Away Team") +  
  theme_minimal() +  
  theme(
    text = element_text(size = 12),  
    axis.text.x = element_text(angle = 45, hjust = 1),  
    legend.position = "right"  
  ) +
  scale_color_manual(values = c("Arsenal" = "red", 
                                "Leicester" = "blue", 
                                "Tottenham" = "black",
                                "Man United" = "green",
                                "Man City" = "purple"))

premier_league_full <- rbind(premier_league_12_15h, premier_league_12_15away)

view(premier_league_full)

colnames(premier_league_12_15h)[2] <- "Team"

colnames(premier_league_12_15away)[2] <- "Team"


prem_full <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ Team + Year, 
                       data = premier_league_full, 
                       FUN = mean)

ggplot(prem_full, aes(x = Year)) + 
  geom_line(aes(y = HST, color = "Shots"), size = 1) + 
  geom_line(aes(y = FTHG, color = "Goals"), size = 1) + 
  facet_wrap(~ Team) +  # Facet by Team
  labs(title = "Home Games: Average Shots and Goals Scored Over Years",
       y = "Average Value", 
       color = "Metric") +
  theme_minimal()

```



```{r}
## aggregating away stats from previous years for comparison with 2015/16 season
avg_away_stats14 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ AwayTeam, 
                              data = premier_league_14, 
                              FUN = mean)

avg_away_stats13 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ AwayTeam, 
                              data = premier_league_13, 
                              FUN = mean)

avg_away_stats12 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ AwayTeam, 
                              data = premier_league_12, 
                              FUN = mean)


premier_league_home15a <-  rbind(avg_away_stats[avg_away_stats$AwayTeam == "Man City", ], 
                                 avg_away_stats[avg_away_stats$AwayTeam == "Man United", ], 
                                 avg_away_stats[avg_away_stats$AwayTeam == "Chelsea", ], 
                                 avg_away_stats[avg_away_stats$AwayTeam == "Leicester", ])

premier_league_home14a <-  rbind(avg_away_stats14[avg_away_stats14$AwayTeam == "Man City", ], 
                                 avg_away_stats14[avg_away_stats14$AwayTeam == "Man United", ], 
                                 avg_away_stats14[avg_away_stats14$AwayTeam == "Chelsea", ], 
                                 avg_away_stats14[avg_away_stats14$AwayTeam == "Leicester", ])

premier_league_home13a <-  rbind(avg_away_stats13[avg_away_stats13$AwayTeam == "Man City", ], 
                                 avg_away_stats13[avg_away_stats13$AwayTeam == "Man United", ], 
                                 avg_away_stats13[avg_away_stats13$AwayTeam == "Chelsea", ], 
                                 avg_away_stats13[avg_away_stats13$AwayTeam == "Leicester", ])

premier_league_home12a <-    rbind(avg_away_stats12[avg_away_stats12$AwayTeam == "Man United", ], 
                                   avg_away_stats12[avg_away_stats12$AwayTeam == "Man City", ], 
                                   avg_away_stats12[avg_away_stats12$AwayTeam == "Chelsea", ], 
                                   avg_away_stats12[avg_away_stats12$AwayTeam == "Leicester", ])

Year <- rep(2015,4)
premier_league_home15a <- cbind(Year, premier_league_home15a)

Year <- rep(2014,4)
premier_league_home14a <- cbind(Year, premier_league_home14a)

Year <- rep(2013,3)
premier_league_home13a <- cbind(Year, premier_league_home13a)

Year <- rep(2012,3)
premier_league_home12a <- cbind(Year, premier_league_home12a)


premier_league_12_15a <- rbind(premier_league_home15a, premier_league_home14a, premier_league_home13a, premier_league_home12a)

View(premier_league_12_15a)



##making graphs
ggplot(premier_league_12_15a, aes(x = Year)) + 
  geom_line(aes(y = AS, color = "Shots"), size = 1) + 
  geom_line(aes(y = FTAG, color = "Goals"), size = 1) + 
  facet_wrap(~ AwayTeam) +  # Facet by Team
  labs(title = "Away Games: Average Shots and Goals Scored Over Years",
       y = "Average Value", 
       color = "Metric") +
  theme_minimal()


ggplot(premier_league_12_15a, aes(x = Year)) + 
  geom_line(aes(y = HS, color = "Shots"), size = 1) + 
  geom_line(aes(y = FTHG, color = "Goals"), size = 1) + 
  facet_wrap(~ AwayTeam) +  # Facet by Team
  labs(title = "Away Games: Average Shots and Goals Conceded Over Years",
       y = "Average Value", 
       color = "Metric") +
  theme_minimal() 


premier_league_12_15away <- premier_league_12_15a[premier_league_12_15a$Year == "2015" | premier_league_12_15a$Year == "2014", ]

premier_league_12_15away$Year <- as.Date(premier_league_12_15away$Year, format = "%Y")

ggplot(premier_league_12_15away, aes(x = Year)) + 
  geom_line(aes(y = AS, color = "Shots"), size = 1) + 
  geom_line(aes(y = FTAG, color = "Goals"), size = 1) + 
  facet_wrap(~ AwayTeam) +  # Facet by Team
  labs(title = "Away Games: Average Shots and Goals Scored Over Years",
       y = "Average Value", 
       color = "Metric") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2014, 2015, by = 1))



full_home_stats <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ HomeTeam, 
                             data = full_premier_league, 
                             FUN = mean)

## aggregating data from 2014/15
avg_home_stats14 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ HomeTeam, 
                              data = premier_league1, 
                              FUN = mean)

avg_away_stats14 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ AwayTeam, 
                              data = premier_league1, 
                              FUN = mean)


##aggregating previous seasons for comparison with the 2015/16 season

avg_home_stats14 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ HomeTeam, 
                              data = premier_league_14, 
                              FUN = mean)


avg_home_stats13 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ HomeTeam, 
                              data = premier_league_13, 
                              FUN = mean)

avg_home_stats12 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ HomeTeam, 
                              data = premier_league_12, 
                              FUN = mean)

#making a new data frame to compare teams of interest
premier_league_home15 <-  rbind(avg_home_stats[avg_home_stats$HomeTeam == "Man City", ], 
                                avg_home_stats[avg_home_stats$HomeTeam == "Man United", ], 
                                avg_home_stats[avg_home_stats$HomeTeam == "Chelsea", ], 
                                avg_home_stats[avg_home_stats$HomeTeam == "Leicester", ])

premier_league_home14 <-  rbind(avg_home_stats14[avg_home_stats14$HomeTeam == "Man City", ], 
                                avg_home_stats14[avg_home_stats14$HomeTeam == "Man United", ], 
                                avg_home_stats14[avg_home_stats14$HomeTeam == "Chelsea", ], 
                                avg_home_stats14[avg_home_stats14$HomeTeam == "Leicester", ])

premier_league_home13 <-  rbind(avg_home_stats13[avg_home_stats13$HomeTeam == "Man City", ], 
                                avg_home_stats13[avg_home_stats13$HomeTeam == "Man United", ], 
                                avg_home_stats13[avg_home_stats13$HomeTeam == "Chelsea", ], 
                                avg_home_stats13[avg_home_stats13$HomeTeam == "Leicester", ])

premier_league_home12 <-    rbind(avg_home_stats12[avg_home_stats12$HomeTeam == "Man United", ], 
                                  avg_home_stats12[avg_home_stats12$HomeTeam == "Man City", ], 
                                  avg_home_stats12[avg_home_stats12$HomeTeam == "Chelsea", ], 
                                  avg_home_stats12[avg_home_stats12$HomeTeam == "Leicester", ])


Year <- rep(2015,4)
premier_league_home15 <- cbind(Year, premier_league_home15)

Year <- rep(2014,4)
premier_league_home14 <- cbind(Year, premier_league_home14)

Year <- rep(2013,3)
premier_league_home13 <- cbind(Year, premier_league_home13)

Year <- rep(2012,3)
premier_league_home12 <- cbind(Year, premier_league_home12)

premier_league_12_15 <- rbind(premier_league_home15, premier_league_home14, premier_league_home13, premier_league_home12)

str(premier_league_12_15)

str(premier_league)

ggplot(premier_league_12_15, aes(Year, HST, color=HomeTeam)) +
  geom_point() +
  geom_line() +
  theme_minimal()

ggplot(premier_league_12_15, aes(FTAG, AST, color=HomeTeam)) +
  geom_point() +
  geom_line() +
  theme_minimal()

premier_league_12_15h <- premier_league_12_15[premier_league_12_15$Year == "2015" | premier_league_12_15$Year == "2014", ]

# Line Plot for Shots and Goals Scored with faceting by Team
ggplot(premier_league_12_15, aes(x = Year)) + 
  geom_line(aes(y = HST, color = "Shots"), size = 1) + 
  geom_line(aes(y = FTHG, color = "Goals"), size = 1) + 
  facet_wrap(~ HomeTeam) +  # Facet by Team
  labs(title = "Home Games: Average Shots and Goals Scored Over Years",
       y = "Average Value", 
       color = "Metric") +
  theme_minimal()


# Line Plot for Shots and Goals with faceting by Team
ggplot(premier_league_12_15, aes(x = Year)) + 
  geom_line(aes(y = AS, color = "Shots"), size = 1) + 
  geom_line(aes(y = FTAG, color = "Goals"), size = 1) + 
  facet_wrap(~ HomeTeam) +  # Facet by Team
  labs(title = "Home Games: Average Shots and Goals Conceded Over Years",
       y = "Average Value", 
       color = "Metric") +
  theme_minimal()


ggplot(premier_league_12_15h, aes(x = Year)) + 
  geom_line(aes(y = AS, color = "Shots"), size = 1) + 
  geom_line(aes(y = FTAG, color = "Goals"), size = 1) + 
  facet_wrap(~ HomeTeam) +  # Facet by Team
  labs(title = "Home Games: Average Shots and Goals Conceded Over Years",
       y = "Average Value", 
       color = "Metric") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2014, 2015, by = 1))

## aggregating data from 2014/15
avg_home_stats14 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ HomeTeam, 
                              data = premier_league1, 
                              FUN = mean)

avg_away_stats14 <- aggregate(cbind(FTHG, FTAG, HS, AS, HST, AST, HF, AF, HC, AC, HY, AY, HR, AR) ~ AwayTeam, 
                              data = premier_league1, 
                              FUN = mean)
```

